A Simple Type Theory for Category Theory

### 2022-04-25 paper/paper.tex
intro tweaks

More experienced users of category theory in programming language
theory find that for modeling computational effects such as general
recursion or indexed-state, or modeling logics for parametricity,
gradual typing, or step-indexed logical relations involves not just
ordinary category theory, but more general notions such as enriched,
internal or multi-categories. However, model construction theorems and
especially soundness and completeness theorems in these areas will
require their own, even more complex versions of the same common
lemmas of category theory, in addition to the domain-specific results
such as existence of fixed points.

At this level of generality, the proofs can be difficult to
understand, as they often involve some combination of the original
category theoretic proof with some additional side-conditions that
everythinng is in fact appropriately step-indexed, continuous etc.
%
One approach to taming this complexity is to prove these general
theorems at a higher level of generality, for instance, proving them
in some abstract 2-category of enriched/internal categories, functors
and transformations.
%
For proofs in generalized category theory specifically, the notion of
a \emph{virtual equipment} has been shown to be occupy an algebraic
sweet-spot of being complex enough to express most basic constructions
(functors, profunctors, natural transformations, adjunctions, kan
extensions, etc) but also general enough to encompass internal,
enriched and ``generalized multi-categories'', which include ordinary
flavors of multi-categories that model intuitionistic and
substructural type theories, but also a great deal more, including
double categories, metric spaces and topological spaces.
%
However, there are two difficulties with this approach.
%
First, the abstract arguments can be difficult to understand compapred
to the original argument for ordinary category theory.
%
Second, the algebraic structure of a virtual equipment is quite
unwieldy to manipulate. The abstraction of a generalized natural
transformation is a kind of ``multi-2-cell'' with many inputs and a
single output. In diagrammatic form these are rendered as

@@

We solve both of these issues by developing an alternative \emph{type
theoretic syntax} for the constructions in a virtual equipment.
%
In this syntax the ``multi-2-cell'' above instead becomes a natural
transformation judgment
\[ \Phi \vdash t : S[F(x);G(y)] \]
where the composable string of input arrows is modeled by an ordered,
linear context $\Phi$ and output $S$ and frames $F,G$ are combined in
the output type.
%
We write natural transformations in a natural deduction style: a
transformation is written as a ``parameterized'' element of the output
profunctor. Furthermore the restriction operators for vertical arrows
into horizontal arrows is naturally modeled by a syntactic
substitution.
%
This interpretation of restriction as substitution both matches the
semantic intuition for the restriction operation as well as greatly
simplifies reasoning about composition of transformations.

>>>>

More experienced practitioners of category theory in programming
language theory find that for modeling computational effects such as
general recursion or indexed-state, or modeling logics for
parametricity, gradual typing, or step-indexed logical relations
involves not just ordinary category theory, but more general notions
such as enriched, internal or multi-categories. However, model
construction theorems and especially soundness and completeness
theorems in these areas will require their own, even more complex
versions of the same common lemmas of category theory, in addition to
the domain-specific results such as existence of fixed points.

At this level of generality, the proofs can be difficult to
understand, as they often involve some combination of the original
category theoretic proof with some additional side-conditions that
everythinng is in fact appropriately step-indexed, continuous etc.
%
One approach to taming this complexity is to prove these general
theorems at a higher level of generality, for instance, proving them
in some abstract 2-category of enriched/internal categories, functors
and transformations.
%
For proofs in generalized category theory specifically, the notion of
a \emph{virtual equipment} has been shown to be occupy an algebraic
sweet-spot of being complex enough to express most basic constructions
(functors, profunctors, natural transformations, adjunctions, kan
extensions, etc) but also general enough to encompass internal,
enriched and ``generalized multi-categories'', which include ordinary
flavors of multi-categories that model intuitionistic and
substructural type theories, but also a great deal more, including
double categories, metric spaces and topological spaces.
%
However, there are two difficulties with this approach.
%
First, the algebraic arguments can be difficult to understand compared
to the original argument for ordinary category theory, which is an
obstacle to wider adoption of this powerful approach.
%
Second, the algebraic structure of a virtual equipment is quite
unwieldy to manipulate. The abstraction of a generalized natural
transformation is a kind of ``multi-2-cell'' with many inputs and a
single output. In diagrammatic form these are rendered as

@@

We address both of these issues by developing an alternative
\emph{type theoretic syntax} for virtual equipments.
%
In this syntax the ``multi-2-cell'' above instead becomes a natural
transformation judgment
\[ \Phi \vdash t : S[F(x);G(y)] \]
where the composable string of input arrows is modeled by an ordered,
linear context $\Phi$ and output $S$ and frames $F,G$ are combined in
the output type using the restriction operation of an equipment.
%
We write natural transformations in a natural deduction style: a
transformation is written as a ``parameterized'' element of the output
profunctor. Furthermore the restriction operators for vertical arrows
into horizontal arrows is naturally modeled by a syntactic
substitution.
%
This interpretation of restriction as substitution both matches the
semantic intuition for the restriction operation as well as greatly
simplifies reasoning about composition of transformations.
